---
title: "Multi-Omics Integration - Maisie Varcoe"
output: html_document
date: "2026-03-10"
---

### Dependencies

```{r}
#if (!require("BiocManager", quietly = TRUE))
 #   install.packages("BiocManager")
#BiocManager::install("MOFA2")
```

```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(patchwork)
library(phyloseq)
library(vegan)
library(tibble)
library(MOFA2)
library(ggrepel)
library(pheatmap)
```

### Load Data

```{r}
# ------------ Microbiome Data -------------------
# Microbiome data: species-level abundance table (already filtered, clean row names)
microbiome <- read.csv(
  "~/Desktop/iawards/metaphlan_output/species_abundance_table_phyloseq/species_table.csv",
  row.names = 1
)
# Load phyloseq object
load("~/Desktop/iawards/metaphlan_output/species_abundance_table_phyloseq/physeq.RData")

# ------------ Metabolomics Data -----------------
# NMR peak area metadata, includes NMR sample IDs and peak area values for metabolites
load("~/Desktop/iawards/NMR metabolomics/NMR_peakarea_metadata.RData")
NMR_meta <- NMR_peakarea_metadata

#--------- Pathway Data ---------------------
pathways <- read.csv("~/Desktop/Research_Project/Scripts/Integration/Data/pathabundance_clean.csv")
# If it's already loaded as an object, check its structure

# ------------ Linking Files -------------------
# Sequencing library - links P22757 IDs to sample numbers
lib <- read_xlsx("~/Desktop/iawards/shotgun_jose_analysis/lib.xlsx")

# Shotgun sample metadata - eg_id, site, lesional, local EASI
sample_meta <- read_xlsx("~/Desktop/iawards/shotgun_jose_analysis/iawards_shotgun_sample_metadata.xlsx")

# Shotgun patient metadata - age, sex, EASI, treatments
shotgun_patient <- read_xlsx("~/Desktop/iawards/shotgun_jose_analysis/iawards_shotgun_patient_metadata.xlsx")

# NMR-shotgun linking file
nmr_shotgun <- read.csv("~/Desktop/iawards/shotgun_jose_analysis/Abundance_tables/NMR_shotgun_sample_names.csv")

# Sample metadata with eg_id linking factor
load("~/Desktop/iawards/metadata/sample_metadata_201106.RData")
link <- sample_metadata_201106

```

# Tier 1: Microbiome & Metabolomics

## Step 1: Data Preparation

Using cleaned data files, extract S. aureus abundance from metaphlan.

Build master linking table, connecting microbiome (SA abundance) to metabolomics via eg_id.

### Data Cleaning

Row 13 (P22757_129, GYECZ1193) was incorrectly assigned GYECZ1197.BACK.LES as its NMR ID. This means:

-   **The duplicate**: BACK.LES appears twice- once correctly (row 22, the real GYECZ1197 sample) and once incorrectly (row 13, the mislinked GYECZ1193 sample)

-   **The "missing" value**: there's no actual NMR data for GYECZ1193 BACK, so the row shouldn't exist at all

Dropping row 13 solves both in one step 

```{r}
# Remove GYECZ1193 BACK - no NMR data available, incorrectly linked to GYECZ1197
nmr_shotgun <- nmr_shotgun[nmr_shotgun$sample_number != "29", ]

# Remove _S1, _S2, etc. suffixes from pathway IDs
pathways$sample_short <- gsub("_S[0-9]+$", "", pathways$sample)
```

### 1.1. Extract SA Abundance from Microbiome Data

```{r}
#------ Check for rows that have S. aureus ---------
sa_rows <- grep("aureus", rownames(microbiome), 
                value=TRUE, ignore.case=TRUE)
print(sa_rows)

# ------ Extract Species-Level SA Row-------
# (not strain level)
sa_abundance <- microbiome["s__Staphylococcus_aureus", ]

# Transpose to dataframe
sa_df <- data.frame(
  sample_id = colnames(microbiome),
  sa_abundance = as.numeric(sa_abundance)
)

# Trim sample IDs to match lib format (remove _S1_L001d_metagenome suffix)
sa_df$sample_short <- gsub("_S[0-9]+_L001d_metagenome", "", 
                            sa_df$sample_id)

print(sa_df)
```

### 1.2. Merge SA Data with Sequencing Data

Joining sa_df to lib because sa_df has IDs (e.g. PXXXXX_101 with abundance X), but doesn't have patient information. So after merging with lib, it will know that patient 1 has abundance X (intermediate ID for linking to later files).

```{r}
# Join to lib to get sample_number
sa_linked <- sa_df %>%
  left_join(
    lib %>% select(Sample, `Submitted Name`),
    by = c("sample_short" = "Sample")
  ) %>%
  rename(sample_number = `Submitted Name`)

# Check - do all 20 rows have a sample_number?
print(sa_linked %>% select(sample_short, sample_number, sa_abundance))
# Check for NAs
sum(is.na(sa_linked$sample_number)) # should be 0
```

### 1.3. Join sa_linked with Sample Metadata

So SA_linked now has PXXXX IDs, sample numbers and SA abundance.

Link this to sample metadata (has sample number, eg_id, sample site, lesional status, local EASI score, S. aureus growth viapath)

Need to link these so that the object with SA abundance now has info on who the patient is and info about them, specifically important that the object now has eg_id which is the linking factor between all data modalities.

```{r}
# Convert sample_meta sample_number to character to match sa_linked
sample_meta$sample_number <- as.character(sample_meta$sample_number)

# Now run the join
sa_linked2 <- sa_linked %>%
  left_join(
    sample_meta %>% 
      select(sample_number, eg_id, sample_site,
             lesional, local_EASI_intensity,
             s_aureus_culture_growth_viapath),
    by = "sample_number"
  )

# Check - do all 20 rows have an eg_id?
print(sa_linked2 %>% select(sample_number, eg_id, 
                              sample_site, lesional, 
                              sa_abundance))
sum(is.na(sa_linked2$eg_id)) # should be 0
```

### 1.4. Join sa_linked2 to Patient Metadata

So the sa_linked2 now has SA abundance, eg_id, sample site and lesional status.

The shotgun_patient has patient-level clinical metadata (age, sex, EASI, weeks since last systemic, weeks since last TCS or TCI, days since any topical, TEWL mean, pH mean).

Need to link these for checking for confounders later on (e.g. age/treatment) and for downstream analyses (EASI = lesionality outcome variable).

```{r}
# Join to get patient-level clinical variables
sa_linked3 <- sa_linked2 %>%
  left_join(
    shotgun_patient %>%
      select(eg_id, age, sex, EASI,
             weeks_since_last_systemic,
             weeks_since_last_TCS_or_TCI,
             days_since_any_topicals,
             TEWL_mean, pH_mean),
    by = "eg_id"
  )

# Check
print(sa_linked3 %>% select(eg_id, sample_site, lesional,
                              sa_abundance, age, sex, EASI))
sum(is.na(sa_linked3$eg_id))
```

### 1.5. Add NMR (Metabolite) to Microbiome Data

Link each microbiome sample to its corresponding NMR sample to create single table where each row has (SA abundance, patient ID, body, site, lesional status, clinical variables, and matching metabolomics ID).

This creates the master linking table which is the foundation for all downstream analysis in Tier 1.

```{r}
# Fix class mismatches for joins
sample_meta$sample_number <- as.character(sample_meta$sample_number)
nmr_shotgun$sample_number <- as.character(nmr_shotgun$sample_number)

#------------ Build master linking table-----------------
master_table <- sa_df %>%
  # Link P22757 IDs to sample numbers via sequencing library
  left_join(lib %>% 
              select(Sample, `Submitted Name`),
            by = c("sample_short" = "Sample")) %>%
  rename(sample_number = `Submitted Name`) %>%
  # Add sample-level metadata (eg_id, site, lesional status)
  left_join(sample_meta %>%
              select(sample_number, eg_id, sample_site,
                     lesional, local_EASI_intensity,
                     s_aureus_culture_growth_viapath),
            by = "sample_number") %>%
  # Add patient-level clinical variables
  left_join(shotgun_patient %>%
              select(eg_id, age, sex, EASI,
                     weeks_since_last_systemic,
                     weeks_since_last_TCS_or_TCI,
                     days_since_any_topicals,
                     TEWL_mean, pH_mean),
            by = "eg_id") %>%
  # Add NMR metabolomics sample IDs
  left_join(nmr_shotgun %>%
              select(sample_number, NMR_sample_id),
            by = "sample_number") %>%
  # Add pathways 
  left_join(pathways %>% select(-sample),
            by = c("sample_short" = "sample_short"))

master_table
dim(master_table)                          
length(unique(master_table$eg_id))         
table(master_table$lesional)               
summary(master_table$sa_abundance)         
colSums(is.na(master_table))               
```

## Step 2: Confounder Checking

Plotting S. aureus abundance against age, EASI score, treatment washout, TWEL and pH across 9 patients. Seeing if any of the variables correlate with SA abundance.

Spearman correlations.

Flag anything that should be carried as a covariate downstream.

### 2.1. Aggregate to Patient Level

Some samples are from the same patient, for these, merge the clinical variables to just one repeat per patient.

```{r}
# One row per patient for visualisation only
patient_level <- master_table %>%
  group_by(eg_id) %>%
  summarise(
    mean_sa = mean(sa_abundance),
    age = first(age),
    sex = first(sex),
    EASI = first(EASI),
    weeks_since_last_systemic = first(weeks_since_last_systemic),
    weeks_since_last_TCS_or_TCI = first(weeks_since_last_TCS_or_TCI),
    days_since_any_topicals = first(days_since_any_topicals),
    TEWL_mean = first(TEWL_mean),
    pH_mean = first(pH_mean)
  )

print(patient_level)

# -------- Clean Systemic Treatment -------------
# Systemic treatment column is numeric and text -> convert to binary
# Convert weeks_since_last_systemic to binary
# never = no recent systemic treatment
# anything else = had systemic treatment
patient_level <- patient_level %>%
  mutate(had_systemic = ifelse(
    weeks_since_last_systemic == "never", 
    "never", "yes"))

print(patient_level %>% select(eg_id, mean_sa, had_systemic))
```

### 2.2. Confounder Plots

Plotting variables to see if any associate with S. aureus levels.

Using scatter plots with spearman R value displayed on plots via stat_cor for visual correlation.

```{r}
patient_level <- patient_level %>%
  mutate(
    TEWL_mean = as.numeric(ifelse(TEWL_mean == "none", NA, TEWL_mean)),
    pH_mean = as.numeric(ifelse(pH_mean == "none", NA, pH_mean)),
    days_since_any_topicals = as.numeric(ifelse(days_since_any_topicals == "none", NA, days_since_any_topicals))
  )

# Plot theme
my_theme <-
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
# ---- Continuous confounders ----
# Age vs SA
age_SA <- ggplot(patient_level, aes(x = age, y = mean_sa)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, colour = "steelblue") +
  stat_cor(method = "spearman", label.x.npc = "left", size=6) +
  labs(title = "Age vs SA Abundance",
       x = "Age", y = "Mean SA Abundance (%)") +
  my_theme

# EASI vs SA
EASI_SA <- ggplot(patient_level, aes(x = EASI, y = mean_sa)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, colour = "steelblue") +
  stat_cor(method = "spearman", label.x.npc = "left", size=6) +
  labs(title = "EASI vs SA Abundance",
       x = "EASI Score", y = "Mean SA Abundance (%)") +
  my_theme

# weeks since TCS vs SA
TCA_SA <- ggplot(patient_level, aes(x = weeks_since_last_TCS_or_TCI, y = mean_sa)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, colour = "steelblue") +
  stat_cor(method = "spearman", label.x.npc = "left", size=6) +
  labs(title = "Weeks Since TCS/TCI vs SA Abundance",
       x = "Weeks Since Last TCS/TCI", y = "Mean SA Abundance (%)") +
  my_theme

# days since topicals vs SA
top_SA <- ggplot(patient_level, aes(x = days_since_any_topicals, y = mean_sa)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, colour = "steelblue") +
  stat_cor(method = "spearman", label.x.npc = "left", size=6) +
  labs(title = "Days Since Any Topicals vs SA Abundance",
       x = "Days Since Any Topicals", y = "Mean SA Abundance (%)") +
  my_theme

# TEWL vs SA
TEWL_SA <- ggplot(patient_level, aes(x = TEWL_mean, y = mean_sa)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, colour = "steelblue") +
  stat_cor(method = "spearman", label.x.npc = "left", size=6) +
  labs(title = "TEWL vs SA Abundance",
       x = "TEWL Mean", y = "Mean SA Abundance (%)") +
  my_theme

# pH vs SA
PH_SA <- ggplot(patient_level, aes(x = pH_mean, y = mean_sa)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, colour = "steelblue") +
  stat_cor(method = "spearman", label.x.npc = "left", size=6) +
  labs(title = "pH vs SA Abundance",
       x = "pH Mean", y = "Mean SA Abundance (%)") +
  my_theme

# ---- Categorical confounder ----
# Systemic treatment vs SA
sys_SA <- ggplot(patient_level, aes(x = had_systemic, y = mean_sa)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 3) +
  labs(title = "Systemic Treatment vs SA Abundance",
       x = "Had Systemic Treatment", 
       y = "Mean SA Abundance (%)") +
  my_theme

# ---- Combine all plots ----
(age_SA + EASI_SA + TCA_SA) / (top_SA + TEWL_SA + PH_SA) / (sys_SA + plot_spacer() + plot_spacer())

# Save combined plot
ggsave("~/Desktop/Research_Project/Scripts/Integration/confounder_plots.png", 
       plot = (age_SA + EASI_SA + TCA_SA) / (top_SA + TEWL_SA + PH_SA) / (sys_SA + plot_spacer() + plot_spacer()),
       width = 15, 
       height = 12, 
      
       dpi = 300)

```

### 2.3. Spearman Correlation

```{r}
# Run Spearman correlations for each continuous variable
# Excluding days since topicals (SD=0 so correlation is impossible)
confounders <- c("age", "EASI", "weeks_since_last_TCS_or_TCI", 
                 "TEWL_mean", "pH_mean")

# Run cor.test for each and extract rho and p-value
spearman_results <- lapply(confounders, function(var) {
  test <- cor.test(patient_level$mean_sa, 
                   patient_level[[var]], 
                   method = "spearman",
                   use = "complete.obs")
  data.frame(
    variable = var,
    rho = round(test$estimate, 3),
    p_value = round(test$p.value, 3)
  )
}) %>% bind_rows
rownames(spearman_results) <- NULL

print(spearman_results)

#--------Pathway Composition Confounders (sample-level, multivariate)------------
# Test each confounder against pathway composition using PERMANOVA (needed becuase 248 pathways)
# Multivariate equivalent of Spearman correlations in 2.1
confounders_pw <- c("age", "EASI", "weeks_since_last_TCS_or_TCI",
                    "TEWL_mean", "pH_mean", "sa_abundance", "lesional")

# Remove empty sample
pathway_matched_clean <- pathway_matched[rowSums(pathway_matched) > 0, ]
meta_matched_clean <- meta_matched[rownames(pathway_matched_clean), ]

# Confirm
nrow(pathway_matched_clean)  # expect 17
rownames(pathway_matched_clean)  # confirm which samples remain

# Rerun confounder check
pathway_confounder_results <- lapply(confounders_pw, function(var) {
  df <- data.frame(var = meta_matched_clean[[var]],
                   row.names = rownames(pathway_matched_clean))
  df <- df[complete.cases(df), , drop = FALSE]
  pw <- pathway_matched_clean[rownames(df), ]
  
  set.seed(123)
  test <- adonis2(vegdist(pw, method = "bray") ~ var,
                  data = df, permutations = 999)
  data.frame(
    variable = var,
    R2 = round(test$R2[1], 3),
    p_value = round(test$`Pr(>F)`[1], 3)
  )
}) %>% bind_rows()

rownames(pathway_confounder_results) <- NULL
print(pathway_confounder_results)

# -------metabolites confounder check -------
metabolite_confounder_results <- lapply(confounders_pw, function(var) {
  df <- data.frame(var = meta_matched[[var]],
                   row.names = rownames(nmr_matched))
  df <- df[complete.cases(df), , drop = FALSE]
  nm <- nmr_matched[rownames(df), ]
  
  set.seed(123)
  test <- adonis2(dist(nm, method = "euclidean") ~ var,
                  data = df, permutations = 999)
  data.frame(
    variable = var,
    R2 = round(test$R2[1], 3),
    p_value = round(test$`Pr(>F)`[1], 3)
  )
}) %>% bind_rows()

rownames(metabolite_confounder_results) <- NULL
print(metabolite_confounder_results)
```

```{r}
# Plot metabolite and pathway confounders using PCA reduction
library(ggpubr)

# Shared theme matching your SA confounder plots
my_theme <- theme(
  plot.title = element_text(size = 16, face = "bold"),
  axis.title = element_text(size = 16),
  axis.text = element_text(size = 16),
  axis.text.x = element_text(angle = 45, hjust = 1)
)

#-------- Metabolite PCA --------
metabolite_pca <- prcomp(nmr_matched, scale. = TRUE)
var_explained <- round(summary(metabolite_pca)$importance[2, 1:2] * 100, 1)

meta_plot_df <- data.frame(
  PC1 = metabolite_pca$x[, 1],
  PC2 = metabolite_pca$x[, 2],
  age = meta_matched$age,
  EASI = meta_matched$EASI,
  TEWL_mean = meta_matched$TEWL_mean,
  pH_mean = meta_matched$pH_mean,
  sa_abundance = meta_matched$sa_abundance,
  lesional = meta_matched$lesional
)

# Continuous confounder plots - scatter with trend line
plot_met_continuous <- function(x_var, title) {
  ggplot(meta_plot_df, aes(x = .data[[x_var]], y = PC1)) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", se = TRUE, colour = "steelblue") +
    stat_cor(method = "spearman", label.x.npc = "left", size = 5) +
    labs(title = title,
         x = x_var,
         y = paste0("PC1 (", var_explained[1], "%)")) +
    theme_classic() +
    my_theme
}

# Categorical confounder plot
plot_met_categorical <- function(x_var, title) {
  ggplot(meta_plot_df, aes(x = .data[[x_var]], y = PC1,
                            colour = .data[[x_var]])) +
    geom_boxplot(outlier.shape = NA, colour = "black") +
    geom_jitter(width = 0.1, size = 3) +
    scale_colour_manual(values = c("LES" = "salmon", "NON_LES" = "#008080")) +
    labs(title = title,
         x = x_var,
         y = paste0("PC1 (", var_explained[1], "%)")) +
    theme_classic() +
    my_theme +
    theme(legend.position = "none")
}

m1 <- plot_met_continuous("sa_abundance", "SA Abundance vs Metabolite PC1")
m2 <- plot_met_continuous("EASI", "EASI vs Metabolite PC1")
m3 <- plot_met_continuous("pH_mean", "pH vs Metabolite PC1")
m4 <- plot_met_continuous("TEWL_mean", "TEWL vs Metabolite PC1")
m5 <- plot_met_continuous("age", "Age vs Metabolite PC1")
m6 <- plot_met_categorical("lesional", "Lesional vs Metabolite PC1")

metabolite_confounder_plot <- (m1 + m2 + m3) / (m4 + m5 + m6) +
  plot_annotation(title = "Metabolite Composition by Confounder (PCA)")

print(metabolite_confounder_plot)
ggsave("~/Desktop/Research_Project/Scripts/Integration/metabolite_confounder_pca.png",
       plot = metabolite_confounder_plot, width = 15, height = 10, dpi = 300)

#-------- Pathway PCoA --------
pathway_pcoa <- cmdscale(vegdist(pathway_matched_clean, method = "bray"),
                          k = 2, eig = TRUE)
eig_vals <- pathway_pcoa$eig
pcoa_var <- round(eig_vals[1:2] / sum(eig_vals[eig_vals > 0]) * 100, 1)

pathway_plot_df <- data.frame(
  PC1 = pathway_pcoa$points[, 1],
  PC2 = pathway_pcoa$points[, 2],
  age = meta_matched_clean$age,
  EASI = meta_matched_clean$EASI,
  TEWL_mean = meta_matched_clean$TEWL_mean,
  pH_mean = meta_matched_clean$pH_mean,
  sa_abundance = meta_matched_clean$sa_abundance,
  lesional = meta_matched_clean$lesional
)

plot_pw_continuous <- function(x_var, title) {
  ggplot(pathway_plot_df, aes(x = .data[[x_var]], y = PC1)) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", se = TRUE, colour = "steelblue") +
    stat_cor(method = "spearman", label.x.npc = "left", size = 5) +
    labs(title = title,
         x = x_var,
         y = paste0("PCo1 (", pcoa_var[1], "%)")) +
    theme_classic() +
    my_theme
}

plot_pw_categorical <- function(x_var, title) {
  ggplot(pathway_plot_df, aes(x = .data[[x_var]], y = PC1,
                               colour = .data[[x_var]])) +
    geom_boxplot(outlier.shape = NA, colour = "black") +
    geom_jitter(width = 0.1, size = 3) +
    scale_colour_manual(values = c("LES" = "salmon", "NON_LES" = "#008080")) +
    labs(title = title,
         x = x_var,
         y = paste0("PCo1 (", pcoa_var[1], "%)")) +
    theme_classic() +
    my_theme +
    theme(legend.position = "none")
}

p1 <- plot_pw_continuous("sa_abundance", "SA Abundance vs Pathway PCo1")
p2 <- plot_pw_continuous("EASI", "EASI vs Pathway PCo1")
p3 <- plot_pw_continuous("pH_mean", "pH vs Pathway PCo1")
p4 <- plot_pw_continuous("TEWL_mean", "TEWL vs Pathway PCo1")
p5 <- plot_pw_continuous("age", "Age vs Pathway PCo1")
p6 <- plot_pw_categorical("lesional", "Lesional vs Pathway PCo1")

pathway_confounder_plot <- (p1 + p2 + p3) / (p4 + p5 + p6) +
  plot_annotation(title = "Pathway Composition by Confounder (PCoA)")

print(pathway_confounder_plot)
ggsave("~/Desktop/Research_Project/Scripts/Integration/pathway_confounder_pcoa.png",
       plot = pathway_confounder_plot, width = 15, height = 10, dpi = 300)
```

## Step 3: Confirm S. Aureus Structures the Microbiome Composition

Phyloseq object with samples and taxa that matches the species table.

### 3a.1. Add the metadata and SA high/low grouping

The SA distribution is heavily skewed towards high SA - 16 out of 20 samples have SA above 30%. This means a simple median split at 67% gives you 10 high and 10 low, which is balanced, but the "low" group still has quite high SA values (up to 67%).

I will use the median split for the formal PERMANOVA then use continuous SA abundance in the Mantel test which avoids the binary split problem.

```{r}
# Create metadata dataframe matched to phyloseq sample names

# Match master_table to phyloseq sample names using sample_id
physeq_meta <- master_table %>%
  select(sample_id, eg_id, sample_site, lesional, 
         sa_abundance, EASI, age, sex,
         weeks_since_last_TCS_or_TCI,
         TEWL_mean, pH_mean) %>%
  column_to_rownames("sample_id")

# Add metadata to phyloseq object
sample_data(physeq) <- sample_data(physeq_meta)

# ------- S. Aureus high/low Grouping ---------
# Median split for PERMANOVA
median_sa <- median(sample_data(physeq)$sa_abundance)
cat("Median SA abundance:", median_sa, "\n")

sample_data(physeq)$sa_group <- ifelse(
  sample_data(physeq)$sa_abundance >= median_sa,
  "SA_high", "SA_low")

table(sample_data(physeq)$sa_group)
```

### 3a.2. PERMANOVA

PERMANOVA confirms that the microbiome composition differs by SA high/low.

```{r}
# Extract OTU table and metadata
otu_mat <- as.data.frame(otu_table(physeq))

# Transpose if necessary - vegan needs samples as rows
if(taxa_are_rows(physeq)){
  otu_mat <- t(otu_mat)
}

meta <- data.frame(sample_data(physeq))

# Calculate Bray-Curtis distance matrix
bray_dist <- vegdist(otu_mat, method = "bray")

# PERMANOVA with SA group, blocking by patient
set.seed(123)
permanova_result <- adonis2(
  bray_dist ~ sa_group,
  data = meta,
  permutations = 999,
  strata = meta$eg_id # Accounts for fact that multiple samples come from same patient
)

print(permanova_result)
```

### 3a.3. Mantel Test (with continuous S. aureus variable)

This uses continuous SA rather than binary split to complement the PERMANOVA result. If both agree = robust.

```{r}
# Mantel test - continuous SA vs microbiome distance
# Build SA distance matrix
sa_dist <- dist(meta$sa_abundance)

# Run Mantel test
set.seed(123)
mantel_result <- mantel(bray_dist, sa_dist,
                        method = "spearman",
                        permutations = 999)

print(mantel_result)
```

### 3a.4. Alpha Diversity

Alpha diversity = measuring the diversity within a single sample = how many different species are present and how evenly distributed are they. One number per sample summarises its complexity.

Measuring:

1.  **Observed (richness)** - counts how many different species are present in a sample (doesn't measure abundance)
2.  **Shannon diversity** - accounts for both number of species present and how evenly distributed their abundances are.

Will correlate each sample's alpha diversity score with its SA abundance using Spearman's correlation. (Sig correlation if SA high samples have lower diversity than SA low samples).

Shannon and Richness:

```{r}
# Calculate alpha diversity
# Note: singleton warning expected for MetaPhlAn relative abundance data
# MetaPhlAn does not produce count data so absence of singletons is normal
alpha_div <- estimate_richness(physeq, 
                               measures = c("Shannon", "Observed"))

# Add SA abundance and patient ID
alpha_div$sa_abundance <- meta$sa_abundance
alpha_div$eg_id <- meta$eg_id
alpha_div$sa_group <- meta$sa_group

# Spearman correlations
shannon_cor <- cor.test(alpha_div$sa_abundance, 
                        alpha_div$Shannon,
                        method = "spearman")

richness_cor <- cor.test(alpha_div$sa_abundance,
                         alpha_div$Observed,
                         method = "spearman")

cat("Shannon vs SA - rho:", round(shannon_cor$estimate, 3),
    "p:", format(shannon_cor$p.value, scientific = TRUE, digits = 3), "\n")
cat("Richness vs SA - rho:", round(richness_cor$estimate, 3),
    "p:", format(richness_cor$p.value, scientific = TRUE, digits = 3), "\n")
```

Plot Alpha Diversity:

```{r}
p_shannon <- ggplot(alpha_div, aes(x = sa_abundance, y = Shannon)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, colour = "steelblue") +
  stat_cor(method = "spearman", label.x.npc = "left", size = 5) +
  labs(title = "Shannon Diversity vs SA Abundance",
       x = "SA Abundance (%)", y = "Shannon Diversity") +
  theme_classic() +
  my_theme

p_richness <- ggplot(alpha_div, aes(x = sa_abundance, y = Observed)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, colour = "steelblue") +
  stat_cor(method = "spearman", label.x.npc = "left", size = 5) +
  labs(title = "Species Richness vs SA Abundance",
       x = "SA Abundance (%)", y = "Observed Species") +
  theme_classic() +
  my_theme

p_shannon + p_richness

ggsave("~/Desktop/Research_Project/Scripts/Integration/alpha_diversity_plots.png",
       plot = p_shannon + p_richness,
       width = 12, height = 5, dpi = 300)
```

### 3b. Add Pathway Data

```{r}
#--------Confirm SA Structures Pathway Composition --------

# Add SA group to meta_matched_clean
meta_matched_clean$sa_group <- ifelse(
  meta_matched_clean$sa_abundance >= median(meta_matched_clean$sa_abundance),
  "SA_high", "SA_low")

# PERMANOVA with SA group
set.seed(123)
permanova_pathway <- adonis2(
  vegdist(pathway_matched_clean, method = "bray") ~ sa_group,
  data = meta_matched_clean,
  permutations = 999,
  strata = meta_matched_clean$eg_id
)
print(permanova_pathway)

# Mantel test - continuous SA vs pathway distance
sa_dist_pw <- dist(meta_matched_clean$sa_abundance)
pathway_dist_clean <- vegdist(pathway_matched_clean, method = "bray")

set.seed(123)
mantel_pathway <- mantel(pathway_dist_clean, sa_dist_pw,
                         method = "spearman",
                         permutations = 999)
print(mantel_pathway)

# Pathway richness (number of pathways detected per sample)
pathway_richness <- data.frame(
  sample = rownames(pathway_matched_clean),
  richness = rowSums(pathway_matched_clean > 0),
  sa_abundance = meta_matched_clean$sa_abundance,
  sa_group = meta_matched_clean$sa_group
)

# Spearman correlation
richness_cor_pw <- cor.test(pathway_richness$sa_abundance,
                             pathway_richness$richness,
                             method = "spearman")
cat("Pathway richness vs SA - rho:", round(richness_cor_pw$estimate, 3),
    "p:", format(richness_cor_pw$p.value, scientific = TRUE, digits = 3), "\n")

# Plot
p_pathway_richness <- ggplot(pathway_richness, 
                              aes(x = sa_abundance, y = richness)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, colour = "steelblue") +
  stat_cor(method = "spearman", label.x.npc = "left", size = 5) +
  labs(title = "Pathway Richness vs SA Abundance",
       x = "SA Abundance (%)", 
       y = "Number of Pathways Detected") +
  theme_classic() +
  my_theme

print(p_pathway_richness)

ggsave("~/Desktop/Research_Project/Scripts/Integration/pathway_richness_sa.png",
       plot = p_pathway_richness,
       width = 6, height = 5, dpi = 300)

# -------- pH Sensitivity Checks --------
# Using pathway_matched_clean (17 samples) and meta_matched_clean

library(vegan)

# Ensure pH is numeric
meta_matched_clean$pH_mean <- as.numeric(
  ifelse(meta_matched_clean$pH_mean == "none", NA, meta_matched_clean$pH_mean))

# Remove samples with missing pH
ph_samples <- rownames(meta_matched_clean)[!is.na(meta_matched_clean$pH_mean)]
cat("Samples with pH data:", length(ph_samples), "\n")

pathway_ph <- pathway_matched_clean[ph_samples, ]
meta_ph    <- meta_matched_clean[ph_samples, ]

# Bray-Curtis distance for pH subset
bc_dist_ph <- vegdist(pathway_ph, method = "bray")

# -------- 1. PERMANOVA - pH --------
set.seed(123)
perm_ph <- adonis2(bc_dist_ph ~ pH_mean, 
                   data = meta_ph,
                   permutations = 999)
print(perm_ph)

# Compare to SA PERMANOVA in same subset
set.seed(123)
perm_sa_ph <- adonis2(bc_dist_ph ~ sa_abundance,
                      data = meta_ph,
                      permutations = 999)
print(perm_sa_ph)

# -------- 2. Mantel - pH --------
ph_dist <- dist(meta_ph$pH_mean)
sa_dist <- dist(meta_ph$sa_abundance)

mantel_ph <- mantel(bc_dist_ph, ph_dist, 
                    method = "spearman", permutations = 999)
cat("\nMantel test - pathway distance vs pH:\n")
print(mantel_ph)

mantel_sa_ph <- mantel(bc_dist_ph, sa_dist,
                       method = "spearman", permutations = 999)
cat("\nMantel test - pathway distance vs SA (pH subset):\n")
print(mantel_sa_ph)

# -------- 3. Alpha Diversity - pH --------
pathway_richness_ph <- rowSums(pathway_ph > 0)

cat("\nSpearman: pathway richness vs pH\n")
cor.test(pathway_richness_ph, meta_ph$pH_mean, method = "spearman")

cat("\nSpearman: pathway richness vs SA (pH subset)\n")
cor.test(pathway_richness_ph, meta_ph$sa_abundance, method = "spearman")

# -------- Plot pH vs pathway richness --------
ph_df <- data.frame(
  sample = ph_samples,
  pH_mean = meta_ph$pH_mean,
  sa_abundance = meta_ph$sa_abundance,
  pathway_richness = pathway_richness_ph,
  lesional = meta_ph$lesional
)

p_ph_rich <- ggplot(ph_df, aes(x = pH_mean, y = pathway_richness,
                                colour = lesional)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, colour = "black", linetype = "dashed") +
  scale_colour_manual(values = c("LES" = "#E41A1C", "NON_LES" = "#377EB8")) +
  labs(title = "Pathway Richness vs Skin pH",
       x = "pH", y = "Pathway Richness",
       colour = "Lesional Status") +
  theme_classic()

p_ph_rich

ggsave("~/Desktop/Research_Project/Scripts/Integration/MOFA+/pathway_richness_pH.png",
       plot = p_ph_rich, width = 7, height = 5, dpi = 300, bg = "white")
```

## Step 4: MOFA+

MOFA+ finds latent factors between data groups, allowing you to then correlate these factors with SA abundance or lesionality to see if the data types co-vary.

### 4.1. Prepare and Align Data

```{r}
# Reload with fresh name to avoid any cached state
load("~/Desktop/iawards/NMR metabolomics/NMR_peakarea_metadata.RData")

# Check the column exists before setting rownames
head(colnames(NMR_peakarea_metadata))
head(NMR_peakarea_metadata$NMR_sample_id)

# Notes:
# - GYECZ1193 BACK excluded - no NMR data (corrected in nmr_shotgun at load time)
# - GYECZ0776 and GYECZ1111 excluded - no microbiome data
# - Final aligned dataset: 18 samples

# ------- Prepare NMR metabolomics matrix --------------
# Set sample IDs as row names and clean metabolite names
# Remove zero variance species
rownames(NMR_peakarea_metadata) <- NMR_peakarea_metadata$NMR_sample_id
NMR_peakarea_metadata$NMR_sample_id <- NULL
colnames(NMR_peakarea_metadata) <- gsub("'", "", colnames(NMR_peakarea_metadata))

# --------- Prepare species matrix ---------------
# Transpose so samples are rows, replace P22757 IDs with NMR sample IDs
species_matched <- t(microbiome)
rownames(species_matched) <- master_table$NMR_sample_id[
  match(rownames(species_matched), master_table$sample_id)
]

# --- Align both matrices to shared samples ---
keep_samples <- rownames(species_matched)[
  rownames(species_matched) %in% rownames(NMR_peakarea_metadata)
]

species_matched <- species_matched[keep_samples, ]

# Remove zero variance species after subsetting to final 18 samples
species_matched <- species_matched[, apply(species_matched, 2, var) != 0]

nmr_matched <- NMR_peakarea_metadata[keep_samples, ]

# --------- Prepare pathway matrix ---------------  
pathway_cols <- colnames(pathways)[!colnames(pathways) %in% 
                                     c("sample", "sample_short")]
pathway_matched <- master_table[!is.na(master_table$NMR_sample_id), 
                                 pathway_cols]
rownames(pathway_matched) <- master_table$NMR_sample_id[
  !is.na(master_table$NMR_sample_id)]
pathway_matched <- pathway_matched[keep_samples, ]  # align to same 18 samples

# --------- Prepare meta_matched ----------------- 
meta_matched <- master_table[!is.na(master_table$NMR_sample_id), ] %>%
  mutate(
    TEWL_mean = as.numeric(ifelse(TEWL_mean == "none", NA, TEWL_mean)),
    pH_mean = as.numeric(ifelse(pH_mean == "none", NA, pH_mean)),
    days_since_any_topicals = as.numeric(ifelse(days_since_any_topicals == "none", 
                                                 NA, days_since_any_topicals))
  ) %>%
  filter(!is.na(NMR_sample_id))  # explicitly remove any remaining NA sample IDs

rownames(meta_matched) <- meta_matched$NMR_sample_id
meta_matched <- meta_matched[keep_samples, ]

# Recheck
nrow(meta_matched)
# Verify all matrices
dim(species_matched)                                      # 18 x 44
dim(nmr_matched)                                          # 18 x 122
dim(pathway_matched)                                      # 18 x 248
all(rownames(species_matched) == rownames(pathway_matched)) # TRUE
all(rownames(species_matched) == rownames(meta_matched))    # TRUE
```

### 4.2. MOFA+

```{r}
#------- Create MOFA+ Object ------------------
# Extract metabolite columns only (first 38 columns)
metabolite_cols <- colnames(NMR_peakarea_metadata)[1:38]
nmr_metabolites <- NMR_peakarea_metadata[, metabolite_cols]
nmr_metabolites <- apply(nmr_metabolites, 2, as.numeric)
rownames(nmr_metabolites) <- rownames(NMR_peakarea_metadata)
nmr_matched <- nmr_metabolites[rownames(species_matched), ]

# Remove empty pathway sample (GYECZ1187.RACF.NONLES)
pathway_matched_clean <- pathway_matched[rowSums(pathway_matched) > 0, ]
meta_matched_clean <- meta_matched[rownames(pathway_matched_clean), ]

# Find common samples across all three modalities
common_samples <- Reduce(intersect, list(
  rownames(nmr_matched),
  rownames(species_matched),
  rownames(pathway_matched_clean)
))
cat("Samples in all three modalities:", length(common_samples), "\n") # expect 17

# Subset all matrices to common samples
nmr_matched_mofa     <- nmr_matched[common_samples, ]
species_matched_mofa <- species_matched[common_samples, ]
pathway_matched_mofa <- pathway_matched_clean[common_samples, ]

# Remove zero variance pathways after subsetting
pathway_matched_mofa <- pathway_matched_mofa[,
  apply(pathway_matched_mofa, 2, var) != 0]

# Normalise pathways by total sum scaling
pathway_matched_mofa_norm <- pathway_matched_mofa / rowSums(pathway_matched_mofa)
pathway_matched_mofa_norm <- pathway_matched_mofa_norm[,
  apply(pathway_matched_mofa_norm, 2, var) != 0]

# Feature selection - reduce to top 50 most variable pathways
# Prevents 248 pathway features dominating over 38 metabolites and 44 species
pathway_var <- apply(pathway_matched_mofa_norm, 2, var)
top_pathways <- names(sort(pathway_var, decreasing = TRUE)[1:50])
pathway_matched_mofa_reduced <- pathway_matched_mofa_norm[, top_pathways]

# Verify all matrices
dim(nmr_matched_mofa)              # 17 x 38
dim(species_matched_mofa)          # 17 x 44
dim(pathway_matched_mofa_reduced)  # 17 x 50
all(rownames(nmr_matched_mofa) == rownames(species_matched_mofa))          # TRUE
all(rownames(nmr_matched_mofa) == rownames(pathway_matched_mofa_reduced))  # TRUE

# MOFA+ needs features as rows, samples as columns
mofa_data <- list(
  species     = t(as.matrix(species_matched_mofa)),
  metabolites = t(as.matrix(nmr_matched_mofa)),
  pathways    = t(as.matrix(pathway_matched_mofa_reduced))
)

# Verify dimensions
lapply(mofa_data, dim)
# species:     ~44 x 17
# metabolites: 38  x 17
# pathways:    50  x 17

# Create MOFA object
mofa_obj <- create_mofa(mofa_data)
print(mofa_obj)
plot_data_overview(mofa_obj)

# -------- Train MOFA+ Model --------------
data_opts <- get_default_data_options(mofa_obj)
data_opts$scale_views <- TRUE
model_opts <- get_default_model_options(mofa_obj)
model_opts$num_factors <- 4
train_opts <- get_default_training_options(mofa_obj)
train_opts$seed <- 123

mofa_obj <- prepare_mofa(
  object = mofa_obj,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)

mofa_obj <- run_mofa(mofa_obj, use_basilisk = TRUE)

saveRDS(mofa_obj,
        file = "~/Desktop/Research_Project/Scripts/Integration/mofa_model_3view_top50.rds")

```

```{r}
# ------- Visualisations -----------------
#mofa_obj <- readRDS("~/Desktop/Research_Project/Scripts/Integration/mofa_model.rds")
# Check the trained model
print(mofa_obj)

# Plot variance explained by each factor in each modality
p1 <- plot_variance_explained(mofa_obj)

# Plot total variance explained
p2 <- plot_variance_explained(mofa_obj, plot_total = TRUE)

p1
p2
# Extract factor scores
factors <- get_factors(mofa_obj)[[1]]
dim(factors)  # should be 18 samples x 4 factors
head(factors)

#------ Heatmap of factor scores -----------
# Basic heatmap with pheatmap
factors_heatmap <- pheatmap(factors,
         cluster_rows = TRUE,      # cluster patients
         cluster_cols = FALSE,     # keep Factor1, Factor2... in order
         scale = "column",         # scale each factor to mean=0, sd=1
         color = colorRampPalette(c("blue", "white", "red"))(100),
         main = "MOFA+ Factor Scores per Patient",
         fontsize_row = 8,
         fontsize_col = 10, bg="white")


# Build annotation from sample names
sample_anno <- data.frame(
  Status = ifelse(grepl("NONLES", rownames(factors)), "NONLES", "LES"),
  row.names = rownames(factors)
)

# Define annotation colours
anno_colours <- list(
  Status = c(LES = "#E41A1C", NONLES = "#377EB8")
)

png("~/Desktop/Research_Project/Scripts/Integration/MOFA+/mofa_heatmap3.png", width = 900, height = 700, bg = "white")

pheatmap(factors,
         annotation_row = sample_anno,
         annotation_colors = anno_colours,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         scale = "column",
         color = colorRampPalette(c("#2166AC", "white", "#D6604D"))(100),
         main = "MOFA+ Factor Scores per Patient",
         fontsize = 16,
         fontsize_row = 12,
         fontsize_col = 16,
         cellwidth=80, bg="white")       

dev.off()
# -----------Save plots----------
ggsave("~/Desktop/Research_Project/Scripts/Integration/MOFA+/mofa_variance_total3.png",
       plot = p2,
       width = 6, height = 5, dpi = 300, bg="white")

ggsave("~/Desktop/Research_Project/Scripts/Integration/MOFA+/mofa_variance_per_factor3.png",
       plot = p1,
       width = 6, height = 5, dpi = 300, bg="white")
ggsave("~/Desktop/Research_Project/Scripts/Integration/MOFA+/Factors_per_Patient3.png",
       plot = factors_heatmap,
       width = 6, height = 5, dpi = 300, bg="white")


write.csv(factors, 
          file = "~/Desktop/Research_Project/Scripts/Integration/MOFA+/mofa_factor_scores3.csv",
          row.names = TRUE)


```

### 4.3. Correlate Factor Scores with S. Aureus Abundance

Correlate MOFA+ factor scores with SA abundance to identify which factor tracks with SA.

**Why do this? -\>** MOFA+ factors are hidden variables capturing a pattern of co-variation across data. When factors is correlated with SA abundance and score consistently high it means samples from SA high patients tend to score high on factor X (vice versa for low SA). Which would mean the biological variation captured by factor X tracks with SA abundance. Finding out this would help you answer the biological story of what changes in the microbiome and metabolome when SA is high.

So the Spearman correlation are between each factor's score with SA and EASI (lesionality). For each factor (1-4), across all 18 samples, asking if that samples score on that factor track with how much SA that sample has.

```{r}
# Add SA abundance to factors dataframe
factors_df <- as.data.frame(factors)
factors_df$sa_abundance <- master_table$sa_abundance[
  match(rownames(factors_df), master_table$NMR_sample_id)
]
factors_df$EASI <- master_table$EASI[
  match(rownames(factors_df), master_table$NMR_sample_id)
]
factors_df$lesional <- master_table$lesional[
  match(rownames(factors_df), master_table$NMR_sample_id)
]

# -------- F1&2 Correlations with SA & EASI -------------
# Spearman correlations with SA - Factors 1 and 2 only
sa_cors <- cor(factors_df[, 1:2], 
               factors_df$sa_abundance,
               method = "spearman",
               use = "complete.obs")

print(sa_cors)
# Spearman correlations with EASI - Factors 1 and 2 only
easi_cors <- cor(factors_df[, 1:2],
                 factors_df$EASI,
                 method = "spearman",
                 use = "complete.obs")

print(easi_cors)
# Run cor.test for each factor against SA and EASI
# p values for Factors 1 and 2 only
for(f in paste0("Factor", 1:2)) {
  cat("\n---", f, "---\n")
  
  sa_test <- cor.test(factors_df[[f]], 
                      factors_df$sa_abundance,
                      method = "spearman", exact=FALSE)
  cat("SA:   rho =", round(sa_test$estimate, 3),
      "  p =", round(sa_test$p.value, 3), "\n")
  
  easi_test <- cor.test(factors_df[[f]], 
                        factors_df$EASI,
                        method = "spearman", exact=FALSE)
  cat("EASI: rho =", round(easi_test$estimate, 3),
      "  p =", round(easi_test$p.value, 3), "\n")
}

#------- F1&2 Correlations with all Clincal Variables-------
# Select numeric clinical variables
clinical_vars <- master_table %>%
  select(sa_abundance, EASI, age, TEWL_mean, pH_mean,
         weeks_since_last_TCS_or_TCI) %>%
  mutate(
    TEWL_mean = as.numeric(gsub("none|not_done", NA, TEWL_mean)),
    pH_mean = as.numeric(gsub("none|not_done", NA, pH_mean)),
    EASI = as.numeric(EASI),
    age = as.numeric(age)
  )
# Match to factor scores by NMR sample ID
clinical_vars <- clinical_vars[
  match(rownames(factors_df), master_table$NMR_sample_id), 
]

# Correlate all clinical variables against Factors 1 and 2
for(f in paste0("Factor", 1:2)) {
  cat("\n======", f, "======\n")
  for(v in colnames(clinical_vars)) {
    test <- cor.test(factors_df[[f]], clinical_vars[[v]],
                     method = "spearman", exact = FALSE)
    cat(v, ": rho =", round(test$estimate, 3),
        "  p =", round(test$p.value, 3), "\n")
  }
}

# Add lesional status to factors_df and plot Factor 1 by lesional status
F1_lesional <- ggplot(factors_df, aes(x = lesional, y = Factor1, fill = lesional)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  labs(title = "Factor 1 scores by lesional status") +
  theme_classic()

# Test it statistically
wilcox.test(Factor1 ~ lesional, data = factors_df)

ggsave("~/Desktop/Research_Project/Scripts/Integration/MOFA+/F1_by_Lesionality3.png",
       plot = F1_lesional,
       width = 8, height = 5, dpi = 300) 

# test all factors correlation with lesionality
for(f in paste0("Factor", 1:4)) {
  test <- wilcox.test(factors_df[[f]] ~ factors_df$lesional)
  cat(f, ": p =", round(test$p.value, 3), "\n")
}


```

Exploring the factors

```{r}
plot_variance_explained(mofa_obj, max_r2 =10)


# --- Top Weights ----------
# Plots top n features with the highest absolute loading value
# for a specific factor
# Good for identifying the main drivers of a factor
# high absolute loading value >0.8

target_factors <- 1:4
target_views <- c("species", "metabolites")

for (f in target_factors) {
  for (v in target_views) {
    
    # 1. Create the Top Weights plot
    p <- plot_top_weights(mofa_obj, 
                          view = v, 
                          factor = f, 
                          nfeatures = 10, 
                          scale = TRUE) + 
      ggtitle(paste("Top 10", v, "Weights for Factor", f)) +
      theme(axis.text.y = element_text(size = 10)) # Adjust font here
    
    # 2. Create a dynamic filename
    file_name <- paste0("~/Desktop/Research_Project/Scripts/Integration/MOFA+/top_", v, "_weights_F", f, ".png")
    
    # 3. Save
    ggsave(file_name, plot = p, width = 9, height = 6, dpi = 300)
    
    message(paste("Saved plot for Factor", f, "View", v))
  }
}

```

Three-view model with pathways

```{r}
# ------- Post-hoc Analysis - 3 View Model -------

# Add clinical variables to factors dataframe
factors_df <- as.data.frame(factors)
factors_df$sa_abundance <- master_table$sa_abundance[
  match(rownames(factors_df), master_table$NMR_sample_id)]
factors_df$EASI <- master_table$EASI[
  match(rownames(factors_df), master_table$NMR_sample_id)]
factors_df$lesional <- master_table$lesional[
  match(rownames(factors_df), master_table$NMR_sample_id)]

# -------- Correlations with all factors (not just 1&2) -------------
# Now we know Factors 1 and 2 are SA-associated, test all 4
for(f in paste0("Factor", 1:4)) {
  cat("\n---", f, "---\n")
  
  sa_test <- cor.test(factors_df[[f]],
                      factors_df$sa_abundance,
                      method = "spearman", exact = FALSE)
  cat("SA:   rho =", round(sa_test$estimate, 3),
      "  p =", round(sa_test$p.value, 3), "\n")
  
  easi_test <- cor.test(factors_df[[f]],
                        factors_df$EASI,
                        method = "spearman", exact = FALSE)
  cat("EASI: rho =", round(easi_test$estimate, 3),
      "  p =", round(easi_test$p.value, 3), "\n")
}

# -------- Correlations with all clinical variables -------
clinical_vars <- master_table %>%
  select(sa_abundance, EASI, age, TEWL_mean, pH_mean,
         weeks_since_last_TCS_or_TCI) %>%
  mutate(
    TEWL_mean = as.numeric(gsub("none|not_done", NA, TEWL_mean)),
    pH_mean = as.numeric(gsub("none|not_done", NA, pH_mean)),
    EASI = as.numeric(EASI),
    age = as.numeric(age)
  )
clinical_vars <- clinical_vars[
  match(rownames(factors_df), master_table$NMR_sample_id), ]

# Test all 4 factors against all clinical variables
for(f in paste0("Factor", 1:4)) {
  cat("\n======", f, "======\n")
  for(v in colnames(clinical_vars)) {
    test <- cor.test(factors_df[[f]], clinical_vars[[v]],
                     method = "spearman", exact = FALSE)
    cat(v, ": rho =", round(test$estimate, 3),
        "  p =", round(test$p.value, 3), "\n")
  }
}

# -------- Lesionality -------
F1_lesional <- ggplot(factors_df, aes(x = lesional, y = Factor1, 
                                       fill = lesional)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  scale_fill_manual(values = c("LES" = "#F8766D", "NON_LES" = "#00BFC4")) +
  labs(title = "Factor 1 scores by lesional status") +
  theme_classic()

wilcox.test(Factor1 ~ lesional, data = factors_df)

ggsave("~/Desktop/Research_Project/Scripts/Integration/MOFA+/F1_by_Lesionality_3view.png",
       plot = F1_lesional, width = 8, height = 5, dpi = 300)

# Test all factors against lesionality
for(f in paste0("Factor", 1:4)) {
  test <- wilcox.test(factors_df[[f]] ~ factors_df$lesional)
  cat(f, ": p =", round(test$p.value, 3), "\n")
}

# -------- Variance explained -------
plot_variance_explained(mofa_obj, max_r2 = 10)

# -------- Top weights - now includes pathways view -------
target_factors <- 1:4
target_views <- c("species", "metabolites", "pathways")  # added pathways

for(f in target_factors) {
  for(v in target_views) {
    p <- plot_top_weights(mofa_obj,
                          view = v,
                          factor = f,
                          nfeatures = 10,
                          scale = TRUE) +
      ggtitle(paste("Top 10", v, "Weights for Factor", f)) +
      theme(axis.text.y = element_text(size = 10))
    
    file_name <- paste0("~/Desktop/Research_Project/Scripts/Integration/MOFA+/top_",
                        v, "_weights_F", f, "_3view.png")
    ggsave(file_name, plot = p, width = 9, height = 6, dpi = 300)
    message(paste("Saved plot for Factor", f, "View", v))
  }
}

# Factor 1 scores by SA abundance - scatter plot
factors_df$sa_abundance <- as.numeric(factors_df$sa_abundance)

p_f1_sa <- ggplot(factors_df, aes(x = sa_abundance, y = Factor1, 
                                   colour = lesional)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, colour = "black", linetype = "dashed") +
  scale_colour_manual(values = c("LES" = "#E41A1C", "NON_LES" = "#377EB8")) +
  labs(title = "MOFA+ Factor 1 vs SA Abundance",
       x = "SA Abundance (%)",
       y = "Factor 1 Score",
       colour = "Lesional Status") +
  annotate("text", x = max(factors_df$sa_abundance, na.rm = TRUE) * 0.1,
           y = max(factors_df$Factor1, na.rm = TRUE) * 0.9,
           label = "rho = -0.776, p < 0.001",
           hjust = 0, size = 4.5) +
  theme_classic()

p_f1_sa

ggsave("~/Desktop/Research_Project/Scripts/Integration/MOFA+/F1_by_SA_3view.png",
       plot = p_f1_sa, width = 8, height = 5, dpi = 300, bg = "white")
```

## Step 5: **SA-Metabolite Associations**

-   Linear regression: metabolite \~ SA abundance (as SA abundance increases, do metabolite levels consistently increase or decrease?), adjusted for lesionality.

    -   metabolite = outcome, SA = predictor ("do samples with higher SA abundance have higher or lower levels of each metabolite, independent of whether the skin is lesional?”)

    -   Run separately for every metabolite to get a list of metabolites that significantly associate with SA abundance

    -   Adjusting for lesionality (covariate) – asks whether this relationship changes independent of lesionality (or if the changes are just a result of the lesional skin)

    -   Running hypothesis on all 38 metabolites seperately, each with a significance threshold of p\>0.05, means i could run into the multiple testing problem (false positives)

        -   So use Benjamin-Hochberg correction to control the false discover rate (FDR)

```{r}
#---------SA-Metabolite Linear Regression-------
results <- list()

for(metabolite in colnames(nmr_matched)) {
  df <- data.frame(
    metabolite = as.numeric(nmr_matched[, metabolite]),
    sa_abundance = master_table$sa_abundance[
      match(rownames(nmr_matched), master_table$NMR_sample_id)],
    lesional = master_table$lesional[
      match(rownames(nmr_matched), master_table$NMR_sample_id)]
  )
  model <- lm(metabolite ~ sa_abundance + lesional, data = df)
  results[[metabolite]] <- summary(model)$coefficients["sa_abundance", ]
}

# Combine results into a dataframe
results_df_sa <- do.call(rbind, results)
results_df_sa <- as.data.frame(results_df_sa)
results_df_sa$metabolite <- rownames(results_df_sa)

# Apply BH (FDR) correction
results_df_sa$p_adjusted <- p.adjust(results_df_sa$`Pr(>|t|)`, method = "fdr")

# Rename columns
colnames(results_df_sa) <- c("Estimate", "Std_Error", "t_value", "p_value", "metabolite", "p_adjusted")

# Report nominally significant results (p < 0.05, uncorrected)
sig_nominal_sa <- results_df_sa[results_df_sa$p_value < 0.05, ]
sig_nominal_sa <- sig_nominal_sa[order(sig_nominal_sa$p_value), ]
sig_nominal_sa$metabolite <- NULL
print(sig_nominal_sa)

# ------ Visualise ---------
results_df_sa$significant <- results_df_sa$p_value < 0.05

sa_metabolites <- ggplot(results_df_sa, aes(x = Estimate, y = -log10(p_value), label = metabolite)) +
  geom_point(aes(color = significant), size = 3) +
  geom_label_repel(data = subset(results_df_sa, p_value < 0.05),
                   aes(label = metabolite), size = 3,
                   box.padding = 0.5,
                   point.padding = 0.3,
                   max.overlaps = Inf) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  annotate("text", x = max(results_df_sa$Estimate), y = -log10(0.05) + 0.1,
           label = "p = 0.05", color = "red", hjust = 1, size = 3) +
  scale_color_manual(values = c("grey60", "steelblue"),
                     labels = c("p > 0.05", "p < 0.05"),
                     name = "") +
  labs(title = "S.Aureus-Metabolite Associations",
       x = "Estimate (Effect Size)",
       y = "-log10(p-value)") +
  theme_classic() +
  theme(legend.position = "top")

ggsave("~/Desktop/Research_Project/Scripts/Integration/SA_metabolite_association.png",
       plot = sa_metabolites,
       width = 8, height = 5, dpi = 300)
```

### 5.1. Also looking at Lesionality-Metabolite Correlation

adjusting for SA abundance

```{r}
#---------Lesionality-Metabolite Linear Regression-------
results <- list()

for(metabolite in colnames(nmr_matched)) {
  df <- data.frame(
    metabolite = as.numeric(nmr_matched[, metabolite]),
    sa_abundance = master_table$sa_abundance[
      match(rownames(nmr_matched), master_table$NMR_sample_id)],
    lesional = master_table$lesional[
      match(rownames(nmr_matched), master_table$NMR_sample_id)]
  )
  model <- lm(metabolite ~ lesional + sa_abundance, data = df)
  results[[metabolite]] <- summary(model)$coefficients["lesionalNON_LES", ]
}

# Combine results into a dataframe
results_df_les <- do.call(rbind, results)
results_df_les <- as.data.frame(results_df_les)
results_df_les$metabolite <- rownames(results_df_les)

# Apply BH (FDR) correction
results_df_les$p_adjusted <- p.adjust(results_df_les$`Pr(>|t|)`, method = "fdr")

# Rename columns
colnames(results_df_les) <- c("Estimate", "Std_Error", "t_value", "p_value", "metabolite", "p_adjusted")

# Report nominally significant results (p < 0.05, uncorrected)
sig_nominal_les <- results_df_les[results_df_les$p_value < 0.05, ]
sig_nominal_les <- sig_nominal_les[order(sig_nominal_les$p_value), ]
sig_nominal_les$metabolite <- NULL
print(sig_nominal_les)
```

### 5.2. Also looked at TEWL-Metabolite Correlation

Adjusting for lesion status

```{r}
#---------TEWL-Metabolite Linear Regression-------
results <- list()
master_table$TEWL_mean <- as.numeric(master_table$TEWL_mean)

for(metabolite in colnames(nmr_matched)) {
  df <- data.frame(
    metabolite = as.numeric(nmr_matched[, metabolite]),
    lesional = master_table$lesional[
      match(rownames(nmr_matched), master_table$NMR_sample_id)],
    TEWL_mean = master_table$TEWL_mean[
      match(rownames(nmr_matched), master_table$NMR_sample_id)]
  )
  model <- lm(metabolite ~ TEWL_mean + lesional, data = df)
  results[[metabolite]] <- summary(model)$coefficients["TEWL_mean", ]
}

# Combine results into a dataframe
results_df_tw <- do.call(rbind, results)
results_df_tw <- as.data.frame(results_df_tw)
results_df_tw$metabolite <- rownames(results_df_tw)

# Apply BH (FDR) correction
results_df_tw$p_adjusted <- p.adjust(results_df_tw$`Pr(>|t|)`, method = "fdr")

# Rename columns
colnames(results_df_tw) <- c("Estimate", "Std_Error", "t_value", "p_value", "metabolite", "p_adjusted")

# Report nominally significant results (p < 0.05, uncorrected)
# Note: FDR correction too stringent for n=16 vs 38 metabolites
sig_nominal_tw <- results_df_tw[results_df_tw$p_value < 0.05, ]
sig_nominal_tw <- sig_nominal_tw[order(sig_nominal_tw$p_value), ]
sig_nominal_tw$metabolite <- NULL
print(sig_nominal_tw)

# ------ Visualise ---------
results_df_tw$significant <- results_df_tw$p_value < 0.05

TEWL_metabolites <- ggplot(results_df_tw, aes(x = Estimate, y = -log10(p_value), label = metabolite)) +
  geom_point(aes(color = significant), size = 3) +
  geom_label_repel(data = subset(results_df_tw, p_value < 0.05),
                   aes(label = metabolite), size = 3,
                   box.padding = 0.5,
                   point.padding = 0.3,
                   max.overlaps = Inf) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  annotate("text", x = max(results_df_tw$Estimate), y = -log10(0.05) + 0.1,
           label = "p = 0.05", color = "red", hjust = 1, size = 3) +
  scale_color_manual(values = c("grey60", "steelblue"),
                     labels = c("p > 0.05", "p < 0.05"),
                     name = "") +
  labs(title = "TEWL-Metabolite Associations",
       x = "Estimate (Effect Size)",
       y = "-log10(p-value)") +
  theme_classic() +
  theme(legend.position = "top")

ggsave("~/Desktop/Research_Project/Scripts/Integration/TEWL_metabolite_association.png",
       plot = TEWL_metabolites,
       width = 8, height = 5, dpi = 300)
```

### 5.3. Spearman Correlation

For the SA-metabolite association

```{r}
# Align sa_abundance to nmr_matched row order once
sa_matched <- master_table$sa_abundance[
  match(rownames(nmr_matched), master_table$NMR_sample_id)]

# Run Spearman correlation for each metabolite
spearman_results <- lapply(colnames(nmr_matched), function(metabolite) {
  test <- suppressWarnings(
    cor.test(sa_matched, 
             as.numeric(nmr_matched[, metabolite]), 
             method = "spearman")
  )
  data.frame(
    metabolite = metabolite,
    rho = round(test$estimate, 3),
    p_value = round(test$p.value, 3)
  )
}) %>% bind_rows()

rownames(spearman_results) <- NULL
spearman_results <- spearman_results[order(spearman_results$p_value), ]
print(spearman_results[spearman_results$p_value < 0.05, ])
```

For the lesionality-metabolite association

(ran wilcoxon rank-sum test bc lesionality is binary not continuous)

```{r}
lesional_matched <- master_table$lesional[
  match(rownames(nmr_matched), master_table$NMR_sample_id)]

wilcox_results <- lapply(colnames(nmr_matched), function(metabolite) {
  test <- wilcox.test(as.numeric(nmr_matched[, metabolite]) ~ lesional_matched)
  data.frame(
    metabolite = metabolite,
    p_value = round(test$p.value, 3)
  )
}) %>% bind_rows()

rownames(wilcox_results) <- NULL
wilcox_results <- wilcox_results[order(wilcox_results$p_value), ]
print(wilcox_results[wilcox_results$p_value < 0.05, ])

results_df[results_df$metabolite %in% c("Valine", "Alanine", "Glutamic acid"), ]
```

For the TEWL_metabolite association

```{r}
tewl_matched <- as.numeric(master_table$TEWL_mean[
  match(rownames(nmr_matched), master_table$NMR_sample_id)])

spearman_tewl <- lapply(colnames(nmr_matched), function(metabolite) {
  test <- suppressWarnings(
    cor.test(tewl_matched,
             as.numeric(nmr_matched[, metabolite]),
             method = "spearman")
  )
  data.frame(
    metabolite = metabolite,
    rho = round(test$estimate, 3),
    p_value = round(test$p.value, 3)
  )
}) %>% bind_rows()

rownames(spearman_tewl) <- NULL
spearman_tewl <- spearman_tewl[order(spearman_tewl$p_value), ]
print(spearman_tewl[spearman_tewl$p_value < 0.05, ])

results_df_tw[results_df_tw$metabolite == "Glycerol", ]
```

### 5.4. Visualise Correlations

```{r}
# Create a matrix of p-values for all three analyses
all_metabolites <- unique(c(rownames(sig_nominal_tw), 
                             rownames(sig_nominal_sa),
                             rownames(sig_nominal_les)))

pval_matrix <- data.frame(
  TEWL = results_df_tw$p_value[match(all_metabolites, results_df_tw$metabolite)],
  S.Aureus = results_df_sa$p_value[match(all_metabolites, results_df_sa$metabolite)],
  Lesional_Status = results_df_les$p_value[match(all_metabolites, results_df_les$metabolite)]
)
rownames(pval_matrix) <- all_metabolites

# Create significance star matrix
# Create significance star matrix
star_matrix <- matrix("", nrow = nrow(pval_matrix), ncol = ncol(pval_matrix))
star_matrix[pval_matrix < 0.05]  <- "*"
star_matrix[pval_matrix < 0.01]  <- "**"
star_matrix[pval_matrix < 0.001] <- "***"

breaks <- seq(0, max(-log10(pval_matrix), na.rm = TRUE), length.out = 50)

p <- pheatmap(-log10(pval_matrix), 
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         color = colorRampPalette(c("#f7f7f7", "orange", "darkred"))(50),
         border_color = "white",
         main = "Metabolite Associations (-log10 p-value)",
         angle_col = 45,
         fontsize_row = 10,
         fontsize_col = 11,
         na_col = "lightgrey",
         display_numbers = star_matrix,
         number_color = "black",
         fontsize_number = 10,
         breaks = breaks,
         legend_breaks = c(0, 1.3, 2, 3),
         legend_labels = c("0\n(not significant)", 
                           "1.3\n(p = 0.05)", 
                           "2\n(p = 0.01)", 
                           "3\n(p = 0.001)"))

# Save with caption
png("~/Desktop/Research_Project/Scripts/Integration/metabolite_heatmap.png",
    width = 2400, height = 2200, res = 300)
grid::grid.newpage()

# Add margin at bottom for caption
grid::pushViewport(grid::viewport(y = 0.53, height = 0.92, just = "centre"))
grid::grid.draw(p$gtable)
grid::popViewport()

grid::grid.text("Colour scale: -log10(p-value). Higher values (darker red) indicate stronger significance.\n* p<0.05   ** p<0.01   *** p<0.001",
                x = 0.5, y = 0.04,
                gp = grid::gpar(fontsize = 8, col = "grey40"))
dev.off()
```

## 
