---
title: "Multi-Omics Integration - Maisie Varcoe"
output: html_document
date: "2026-03-10"
---

### Dependencies

```{r}
library(readxl)
library(dplyr)
```

### Load Data

```{r}
# ------------ Microbiome Data -------------------
# Microbiome data
microbiome <- read.delim("~/Desktop/iawards/metaphlan_output/metaphlan_sgb_merged_table.txt",
                         header=TRUE,
                         skip = 1,
                         comment.char = "")

# ------------ Metabolomics Data -----------------
# NMR peak area metadata, includes NMR sample IDs and peak area values for metabolites
load("~/Desktop/iawards/NMR metabolomics/NMR_peakarea_metadata.RData")
NMR_meta <- NMR_peakarea_metadata

# ------------ Linking Files -------------------
# Sequencing library - links P22757 IDs to sample numbers
lib <- read_xlsx("~/Desktop/iawards/shotgun_jose_analysis/lib.xlsx")

# Shotgun sample metadata - eg_id, site, lesional, local EASI
sample_meta <- read_xlsx("~/Desktop/iawards/shotgun_jose_analysis/iawards_shotgun_sample_metadata.xlsx")

# Shotgun patient metadata - age, sex, EASI, treatments
shotgun_patient <- read_xlsx("~/Desktop/iawards/shotgun_jose_analysis/iawards_shotgun_patient_metadata.xlsx")

# NMR-shotgun linking file
nmr_shotgun <- read.csv("~/Desktop/iawards/shotgun_jose_analysis/Abundance_tables/NMR_shotgun_sample_names.csv")

# Sample metadata with eg_id linking factor
load("~/Desktop/iawards/metadata/sample_metadata_201106.RData")
link <- sample_metadata_201106
```

# Tier 1: Microbiome & Metabolomics 

## Step 1: Data Preparation

Using cleaned data files, extract S. aureus abundance from metaphlan.

Build master linking table, connecting microbiome (SA abundance) to metabolomics via eg_id.

```{r}

```

## Step2: Confounder Checking

Plotting S. aureus abundance against age, EASI score, treatment washout, TWEL and pH across 11 patients.

Spearman correlations.

Flag anything that should be carried as a covariate.

```{r}

```
