---
title: "Multi-Omics Integration - Maisie Varcoe"
output: html_document
date: "2026-03-10"
---

### Dependencies

```{r}
library(readxl)
library(dplyr)
```

### Load Data

```{r}
# ------------ Microbiome Data -------------------
# Microbiome data
microbiome <- read.delim("~/Desktop/iawards/metaphlan_output/metaphlan_sgb_merged_table.txt",
                         header=TRUE,
                         skip = 1,
                         comment.char = "")

# ------------ Metabolomics Data -----------------
# NMR peak area metadata, includes NMR sample IDs and peak area values for metabolites
load("~/Desktop/iawards/NMR metabolomics/NMR_peakarea_metadata.RData")
NMR_meta <- NMR_peakarea_metadata

# ------------ Linking Files -------------------
# Sequencing library - links P22757 IDs to sample numbers
lib <- read_xlsx("~/Desktop/iawards/shotgun_jose_analysis/lib.xlsx")

# Shotgun sample metadata - eg_id, site, lesional, local EASI
sample_meta <- read_xlsx("~/Desktop/iawards/shotgun_jose_analysis/iawards_shotgun_sample_metadata.xlsx")

# Shotgun patient metadata - age, sex, EASI, treatments
shotgun_patient <- read_xlsx("~/Desktop/iawards/shotgun_jose_analysis/iawards_shotgun_patient_metadata.xlsx")

# NMR-shotgun linking file
nmr_shotgun <- read.csv("~/Desktop/iawards/shotgun_jose_analysis/Abundance_tables/NMR_shotgun_sample_names.csv")

# Sample metadata with eg_id linking factor
load("~/Desktop/iawards/metadata/sample_metadata_201106.RData")
link <- sample_metadata_201106
```

# Tier 1: Microbiome & Metabolomics 

## Step 1: Data Preparation

Using cleaned data files, extract S. aureus abundance from metaphlan.

Build master linking table, connecting microbiome (SA abundance) to metabolomics via eg_id.

### 1.1. Extract SA Abundance from Microbiome Data

```{r}
# ------ Put taxa names into rownames --------------
# Set the first column as row names
rownames(microbiome) <- microbiome$X
microbiome$X <- NULL

# Verify it worked
head(rownames(microbiome))

#------ Check for rows that have S. aureus ---------
sa_rows <- grep("aureus", rownames(microbiome), 
                value=TRUE, ignore.case=TRUE)
print(sa_rows)

# Check dimensions - should be 176 x 20
dim(microbiome)

# ------ Extract Species-Level SA Row-------
# (not strain level)
sa_row_name <- "k__Bacteria|p__Firmicutes|c__Bacilli|o__Bacillales|f__Staphylococcaceae|g__Staphylococcus|s__Staphylococcus_aureus"

sa_abundance <- microbiome[sa_row_name, ]

# Transpose to dataframe
sa_df <- data.frame(
  sample_id = colnames(microbiome),
  sa_abundance = as.numeric(sa_abundance)
)

# Trim sample IDs to match lib format (remove _S1_L001d_metagenome suffix)
sa_df$sample_short <- gsub("_S[0-9]+_L001d_metagenome", "", 
                            sa_df$sample_id)

print(sa_df)
```

### 1.2. Merge SA Data with Sequencing Data

Joining sa_df to lib because sa_df has IDs (e.g. PXXXXX_101 with abundance X), but doesn't have patient information. So after merging with lib, it will know that patient 1 has abundance X (intermediate ID for linking to later files).

```{r}
# Join to lib to get sample_number
sa_linked <- sa_df %>%
  left_join(
    lib %>% select(Sample, `Submitted Name`),
    by = c("sample_short" = "Sample")
  ) %>%
  rename(sample_number = `Submitted Name`)

# Check - do all 20 rows have a sample_number?
print(sa_linked %>% select(sample_short, sample_number, sa_abundance))
# Check for NAs
sum(is.na(sa_linked$sample_number)) # should be 0
```

### 1.3. Join sa_linked with Sample Metadata

So SA_linked now has PXXXX IDs, sample numbers and SA abundance.

Link this to sample metadata (has sample number, eg_id, sample site, lesional status, local EASI score, S. aureus growth viapath)

Need to link these so that the object with SA abundance now has info on who the patient is and info about them, specifically important that the object now has eg_id which is the linking factor between all data modalities.

```{r}
# Check the classes - one should be numeric and one character
head(sa_linked$sample_number)
head(sample_meta$sample_number)

class(sa_linked$sample_number)
class(sample_meta$sample_number)

# Convert sample_meta sample_number to character to match sa_linked
sample_meta$sample_number <- as.character(sample_meta$sample_number)

# Verify they match now - both should be character
class(sa_linked$sample_number)
class(sample_meta$sample_number)

# Now run the join
sa_linked2 <- sa_linked %>%
  left_join(
    sample_meta %>% 
      select(sample_number, eg_id, sample_site,
             lesional, local_EASI_intensity,
             s_aureus_culture_growth_viapath),
    by = "sample_number"
  )

# Check - do all 20 rows have an eg_id?
print(sa_linked2 %>% select(sample_number, eg_id, 
                              sample_site, lesional, 
                              sa_abundance))
sum(is.na(sa_linked2$eg_id)) # should be 0
```

### 1.4. Join sa_linked2 to Patient Metadata

So the sa_linked2 now has SA abundance, eg_id, sample site and lesional status.

The shotgun_patient has patient-level clinical metadata (age, sex, EASI, weeks since last systemic, weeks since last TCS or TCI, days since any topical, TEWL mean, pH mean).

Need to link these for checking for confounders later on (e.g. age/treatment) and for downstream analyses (EASI = lesionality outcome variable).

```{r}
# Check classes match before joining
class(sa_linked2$eg_id)
class(shotgun_patient$eg_id)

# Join to get patient-level clinical variables
sa_linked3 <- sa_linked2 %>%
  left_join(
    shotgun_patient %>%
      select(eg_id, age, sex, EASI,
             weeks_since_last_systemic,
             weeks_since_last_TCS_or_TCI,
             days_since_any_topicals,
             TEWL_mean, pH_mean),
    by = "eg_id"
  )

# Check
print(sa_linked3 %>% select(eg_id, sample_site, lesional,
                              sa_abundance, age, sex, EASI))
sum(is.na(sa_linked3$eg_id))
```

### 1.5. Add NMR (Metabolite) to Microbiome Data

Link each microbiome sample to its corresponding NMR sample to creat single table where each row has (SA abundance, patient ID, body, site, lesional status, clinical variables, and matching metabolomics ID).

This creates the master linking table which is the foundation for all downstream analysis in Tier 1.

```{r}
# Check classes match
class(sa_linked3$sample_number)
class(nmr_shotgun$sample_number)

# Convert nmr_shotgun sample_number to character
nmr_shotgun$sample_number <- as.character(nmr_shotgun$sample_number)

# Verify
class(sa_linked3$sample_number)
class(nmr_shotgun$sample_number)

# Now run the join
master_table <- sa_linked3 %>%
  left_join(
    nmr_shotgun %>%
      select(sample_number, NMR_sample_id),
    by = "sample_number"
  )

# Final check
print(master_table %>%
  select(eg_id, sample_site, lesional,
         sa_abundance, NMR_sample_id, 
         EASI, age, sex))

sum(is.na(master_table$NMR_sample_id))
```

## Step 2: Confounder Checking

Plotting S. aureus abundance against age, EASI score, treatment washout, TWEL and pH across 11 patients.

Spearman correlations.

Flag anything that should be carried as a covariate.

```{r}

```
