---
title: "Multi-Omics Integration - Maisie Varcoe"
output: html_document
date: "2026-03-10"
---

### Dependencies

```{r}
library(readxl)
library(dplyr)
```

### Load Data

```{r}
# ------------ Microbiome Data -------------------
# Microbiome data: species-level abundance table (already filtered, clean row names)
microbiome <- read.csv(
  "~/Desktop/iawards/metaphlan_output/species_abundance_table_phyloseq/species_table.csv",
  row.names = 1
)

# ------------ Metabolomics Data -----------------
# NMR peak area metadata, includes NMR sample IDs and peak area values for metabolites
load("~/Desktop/iawards/NMR metabolomics/NMR_peakarea_metadata.RData")
NMR_meta <- NMR_peakarea_metadata

# ------------ Linking Files -------------------
# Sequencing library - links P22757 IDs to sample numbers
lib <- read_xlsx("~/Desktop/iawards/shotgun_jose_analysis/lib.xlsx")

# Shotgun sample metadata - eg_id, site, lesional, local EASI
sample_meta <- read_xlsx("~/Desktop/iawards/shotgun_jose_analysis/iawards_shotgun_sample_metadata.xlsx")

# Shotgun patient metadata - age, sex, EASI, treatments
shotgun_patient <- read_xlsx("~/Desktop/iawards/shotgun_jose_analysis/iawards_shotgun_patient_metadata.xlsx")

# NMR-shotgun linking file
nmr_shotgun <- read.csv("~/Desktop/iawards/shotgun_jose_analysis/Abundance_tables/NMR_shotgun_sample_names.csv")

# Sample metadata with eg_id linking factor
load("~/Desktop/iawards/metadata/sample_metadata_201106.RData")
link <- sample_metadata_201106
```

# Tier 1: Microbiome & Metabolomics

## Step 1: Data Preparation

Using cleaned data files, extract S. aureus abundance from metaphlan.

Build master linking table, connecting microbiome (SA abundance) to metabolomics via eg_id.

### 1.1. Extract SA Abundance from Microbiome Data

```{r}
#------ Check for rows that have S. aureus ---------
sa_rows <- grep("aureus", rownames(microbiome), 
                value=TRUE, ignore.case=TRUE)
print(sa_rows)

# ------ Extract Species-Level SA Row-------
# (not strain level)
sa_abundance <- microbiome["s__Staphylococcus_aureus", ]

# Transpose to dataframe
sa_df <- data.frame(
  sample_id = colnames(microbiome),
  sa_abundance = as.numeric(sa_abundance)
)

# Trim sample IDs to match lib format (remove _S1_L001d_metagenome suffix)
sa_df$sample_short <- gsub("_S[0-9]+_L001d_metagenome", "", 
                            sa_df$sample_id)

print(sa_df)
```

### 1.2. Merge SA Data with Sequencing Data

Joining sa_df to lib because sa_df has IDs (e.g. PXXXXX_101 with abundance X), but doesn't have patient information. So after merging with lib, it will know that patient 1 has abundance X (intermediate ID for linking to later files).

```{r}
# Join to lib to get sample_number
sa_linked <- sa_df %>%
  left_join(
    lib %>% select(Sample, `Submitted Name`),
    by = c("sample_short" = "Sample")
  ) %>%
  rename(sample_number = `Submitted Name`)

# Check - do all 20 rows have a sample_number?
print(sa_linked %>% select(sample_short, sample_number, sa_abundance))
# Check for NAs
sum(is.na(sa_linked$sample_number)) # should be 0
```

### 1.3. Join sa_linked with Sample Metadata

So SA_linked now has PXXXX IDs, sample numbers and SA abundance.

Link this to sample metadata (has sample number, eg_id, sample site, lesional status, local EASI score, S. aureus growth viapath)

Need to link these so that the object with SA abundance now has info on who the patient is and info about them, specifically important that the object now has eg_id which is the linking factor between all data modalities.

```{r}
# Convert sample_meta sample_number to character to match sa_linked
sample_meta$sample_number <- as.character(sample_meta$sample_number)

# Now run the join
sa_linked2 <- sa_linked %>%
  left_join(
    sample_meta %>% 
      select(sample_number, eg_id, sample_site,
             lesional, local_EASI_intensity,
             s_aureus_culture_growth_viapath),
    by = "sample_number"
  )

# Check - do all 20 rows have an eg_id?
print(sa_linked2 %>% select(sample_number, eg_id, 
                              sample_site, lesional, 
                              sa_abundance))
sum(is.na(sa_linked2$eg_id)) # should be 0
```

### 1.4. Join sa_linked2 to Patient Metadata

So the sa_linked2 now has SA abundance, eg_id, sample site and lesional status.

The shotgun_patient has patient-level clinical metadata (age, sex, EASI, weeks since last systemic, weeks since last TCS or TCI, days since any topical, TEWL mean, pH mean).

Need to link these for checking for confounders later on (e.g. age/treatment) and for downstream analyses (EASI = lesionality outcome variable).

```{r}
# Join to get patient-level clinical variables
sa_linked3 <- sa_linked2 %>%
  left_join(
    shotgun_patient %>%
      select(eg_id, age, sex, EASI,
             weeks_since_last_systemic,
             weeks_since_last_TCS_or_TCI,
             days_since_any_topicals,
             TEWL_mean, pH_mean),
    by = "eg_id"
  )

# Check
print(sa_linked3 %>% select(eg_id, sample_site, lesional,
                              sa_abundance, age, sex, EASI))
sum(is.na(sa_linked3$eg_id))
```

### 1.5. Add NMR (Metabolite) to Microbiome Data

Link each microbiome sample to its corresponding NMR sample to create single table where each row has (SA abundance, patient ID, body, site, lesional status, clinical variables, and matching metabolomics ID).

This creates the master linking table which is the foundation for all downstream analysis in Tier 1.

```{r}
# Fix class mismatches for joins
sample_meta$sample_number <- as.character(sample_meta$sample_number)
nmr_shotgun$sample_number <- as.character(nmr_shotgun$sample_number)

#------------ Build master linking table-----------------
master_table <- sa_df %>%
  # Link P22757 IDs to sample numbers via sequencing library
  left_join(lib %>% 
              select(Sample, `Submitted Name`),
            by = c("sample_short" = "Sample")) %>%
  rename(sample_number = `Submitted Name`) %>%
  # Add sample-level metadata (eg_id, site, lesional status)
  left_join(sample_meta %>%
              select(sample_number, eg_id, sample_site,
                     lesional, local_EASI_intensity,
                     s_aureus_culture_growth_viapath),
            by = "sample_number") %>%
  # Add patient-level clinical variables
  left_join(shotgun_patient %>%
              select(eg_id, age, sex, EASI,
                     weeks_since_last_systemic,
                     weeks_since_last_TCS_or_TCI,
                     days_since_any_topicals,
                     TEWL_mean, pH_mean),
            by = "eg_id") %>%
  # Add NMR metabolomics sample IDs
  left_join(nmr_shotgun %>%
              select(sample_number, NMR_sample_id),
            by = "sample_number")

master_table
dim(master_table)                          # expect 20 x 18
length(unique(master_table$eg_id))         # expect 9
table(master_table$lesional)               # check LES/NON_LES split
summary(master_table$sa_abundance)         # check 0-100 range
colSums(is.na(master_table))               # check for unexpected NAs
```

## Step 2: Confounder Checking

Plotting S. aureus abundance against age, EASI score, treatment washout, TWEL and pH across 9 patients. Seeing if any of the variables correlate with SA abundance.

Spearman correlations.

Flag anything that should be carried as a covariate downstream.

### Aggregate to Patient Level

Some samples are from the same patient, for these, merge the clinical variables to just one repeat per patient.

```{r}
# One row per patient for visualisation only
patient_level <- master_table %>%
  group_by(eg_id) %>%
  summarise(
    mean_sa = mean(sa_abundance),
    age = first(age),
    sex = first(sex),
    EASI = first(EASI),
    weeks_since_last_systemic = first(weeks_since_last_systemic),
    weeks_since_last_TCS_or_TCI = first(weeks_since_last_TCS_or_TCI),
    days_since_any_topicals = first(days_since_any_topicals),
    TEWL_mean = first(TEWL_mean),
    pH_mean = first(pH_mean)
  )

print(patient_level)

# -------- Clean Systemic Treatment -------------
# Systemic treatment column is numeric and text -> convert to binary
# Convert weeks_since_last_systemic to binary
# never = no recent systemic treatment
# anything else = had systemic treatment
patient_level <- patient_level %>%
  mutate(had_systemic = ifelse(
    weeks_since_last_systemic == "never", 
    "never", "yes"))

print(patient_level %>% select(eg_id, mean_sa, had_systemic))
```
